---
title: ""
output:
  github_document:
    toc: yes
  pdf_document:
    toc: yes
---
 
<center> <h1>Formula 1 Racing: <br /> Does the car or the driver have the greater impact?</h1> </center>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(modelr)

df_timedata <- read_csv("processed_data/std_avg_laptime.csv")
df_data <- read_csv("processed_data/clean_f1.csv")
```

## Data Background

#### Context
The Formula 1 World Championship has been one of the premier forms of auto racing around the world since its inaugural season in 1950. The word "formula" refers to the set of rules to which all participants' cars must conform. A Formula 1 season consists of a series of Grands Prix races which take place worldwide on circuits and closed public roads.

#### Source
The [Formula 1 World Championships](https://www.kaggle.com/rohanrao/formula-1-world-championship-1950-2020) dataset consists of all information on the Formula 1 races, drivers, constructors, qualifying, circuits, lap times, pit stops, and championships for every season from 1950 to 2020.

This dataset was published by Rohan Rao, a Data Scientist who goes by the name [Vopani](https://www.kaggle.com/rohanrao) on Kaggle. (Fun fact: he is also the reigning National Sudoku Champion of India.) He compiled this dataset using the [Ergast Developer API](http://ergast.com/mrd/), an experimental web service that provides a historical record of motor racing data for non-commercial purposes. The API provides data specifically for the Formula One series, from the beginning of the world championships in 1950 up to the present season.

The Ergast API website does not give any details about their data collection procedures, but their data is very thorough and complete. Additionally, Formula 1 world championships are internationally televised and race results are publicly accessible information [(see official Formula 1 website)](https://www.formula1.com/en/results.html), so there is no reason to doubt the accuracy of the data reported.


----------------------------------------

## Investigation Question

#### Does the car or the driver have the greater impact?

Unlike most sports, Formula 1 racing is heavily reliant on the performance capability of their equipment, which is their cars that are created by the constructors. To what extent do the constructor and driver predict race performance, and which one has the greater predictive capability?

----------------------------------------

## Data Tidying

The data came in many separate files, including `results.csv`, `drivers.csv`, `constructors.csv`, `circuits.csv`, `races.csv`, `status.csv`, and `laptimes.csv`. Our first step was to join all of these data frames together by their relevant ID numbers and replace the missing values with `NA`.

In our dataset we kept these columns: 

- resultId
- raceId
- driverId
- constructorId
- positionOrder
- laps
- fastestLapSpeed
- statusId
- driver_name
- constructor_name
- year
- round
- circuitId
- race_name
- status
- circuit_name

```{r}
df_data %>% glimpse()
```

#### Time data

The column `milliseconds` reports the race completion time for all of the drivers that were able to finish the race. However, drivers are often unable to complete the full race due to collisions, car breakdowns, or other problems, leaving many missing values for total race time. We discarded `milliseconds` and created a second dataset where we supplemented information from `laptimes.csv` to compute total race time even when race status was not finished. The lap times were understandably missing when the number of laps completed was 0, so we removed these observations. Overall, we still only have lap times for 9,233 of our 24,900 though. 

In this time-filtered dataset we added these columns: 

- total_time
- avg_lap
- circuit_avg_lap
- circuit_lap_sd
- std_avg_lap

```{r}
df_timedata %>% glimpse()
```

#### Potential Problems

- Some examples of [constructor name changes](https://www.reddit.com/r/formula1/comments/1dos3r/i_made_a_diagram_to_show_how_current_f1_teams/)
- Rules Era can change performance
- New drivers are skewed because they don't have as many data points yet
- This data is for 70 years of races and [formula racing has changed a lot](https://youtu.be/hgLQWIAaCmY)

----------------------------------------

## Exploratory Data Analysis
 
#### Standardize by Average Lap Time

```{r finish race, fig.width=6, fig.height=4, echo=FALSE}
df_data %>% 
  group_by(positionOrder) %>%
  mutate(mean_laps = mean(laps)) %>%
  ungroup() %>%
  ggplot(aes(positionOrder, mean_laps)) +
  geom_point() + 
  labs(
    y = "Mean laps completed",
    x = "Final position ranking"
  ) + 
  theme_minimal() + 
  geom_line(color = "blue")
```

As shown in the graph above, final position ranking is heavily reliant on how many laps the driver was able to complete. And almost a third of the time, collisions or car troubles put drivers out of commission before the race is finished. These final standings are not reflective of how well the driver was doing before that point, so we decided to use average lap time to create a more comprehensive performance metric.

Average lap time is a comprehensive measure of how well a driver performed in a race because it is informed only on the laps they were able to complete, unlike final position. The only problem is that it doesn't account for the effect of circuit on lap time, so we need to correct for this difference across circuits.

```{r average lap time per circuit, echo=FALSE}
df_timedata %>%
  ggplot(aes(fct_reorder(as.factor(circuitId), circuit_avg_lap), circuit_avg_lap)) + 
  geom_point() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  labs(
    x = "Circuit ID number",
    y = "Average lap time (ms)"
  )
```

The average lap time can be drastically different across circuits due to the differences in track length and shape, so we need a way to compare average lap time across circuits. Dividing average lap time by average lap time per circuit doesn't work because it doesn't account for the range of average lap time on each circuit. To correct for the impact of circuit on average lap time, we standardized by circuit:

$$\mu = \sum_{i}^{n} \frac{x_i}{n}$$

$$\sigma = \sqrt{\sum_{i}^{n} \frac{(x_i - \mu)^2}{n}}$$

$$z = \frac{x-\mu}{\sigma}$$

where $x$ is the data, $\mu$ is the mean, $\sigma$ is the standard deviation, and $z$ is the standard score of $x$

## Graphing by standard average lap time

```{r fig.width = 20,fig.height=10, echo=FALSE}
#df_timedata %>%
#  mutate(driver_name = fct_reorder(as.character(driver_name),std_avg_lap))%>%
#  ggplot(mapping = aes(x = driver_name, y = std_avg_lap)) +
#  geom_boxplot() +
#  theme_minimal() +
#  theme(axis.text.x = element_text(angle = 90,vjust = 0.5, hjust = 1))+
#  labs(x = "Driver Name",y = "Standardized Average Lap Time (ms)")

df_timedata %>%
  mutate(constructor_name = fct_reorder(as.character(constructor_name),std_avg_lap, .fun = mean))%>%
  ggplot(mapping = aes(x = constructor_name, y = std_avg_lap)) +
  geom_boxplot()+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90,vjust = 0.5, hjust = 1))+
  labs(x = "Vehicle Constructor Name",y = "Standardized Average Lap Time (ms)")
```

#### Assess Standardized Average Lap Time vs. Circuit

```{r, echo = FALSE}
df_timedata %>%
  ggplot(aes(fct_reorder(as.factor(circuitId), std_avg_lap), std_avg_lap)) +
  geom_boxplot() + 
  labs(
    x = "Circuit ID number",
    y = "standardized average lap time",
    title = "Comparing standardized average lap time by circuit"
  ) + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

The graph above shows a visual comparison of standardized average lap time by circuit. What we hope to see is that standardized average lap time minimizes the effect of circuit, and it seems it does because the slope of the median is small in magnitude relative to the range of the standardized average lap time. In conclusion, standardized average lap time captures overall performance in a way that minimizes the effect of number of laps driven and what circuit the race was held on.

However, when we applied a linear model using standard average lap time as the performance metric with driver and constructor as inputs, we found that it had an rsquare of only 0.08. This discouraging result told us that standardized average lap time is very weakly correlated to driver and constructor, so we decided to take our modelling attempts in another direction.

## Graphing by final position

```{r fig.width = 20,fig.height=10, echo=FALSE}
#df_data %>%
#  mutate(driver_name = fct_reorder(as.character(driver_name),positionOrder))%>%
#  ggplot(mapping = aes(x = driver_name, y = positionOrder)) +
#  geom_boxplot() +
#  theme_minimal() +
#  theme(axis.text.x = element_text(angle = 90,vjust = 0.5, hjust = 1))+
#  labs(x = "Driver Name",y = "Final position ranking")

df_data %>%
  mutate(constructor_name = fct_reorder(as.character(constructor_name),positionOrder, .fun = mean))%>%
  ggplot(mapping = aes(x = constructor_name, y = positionOrder)) +
  geom_boxplot()+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90,vjust = 0.5, hjust = 1))+
  labs(x = "Vehicle Constructor Name",y = "Final position ranking")
```

## Linear Model by final position

\\TODO:
Using final position in the race as our metric of performance, the linear model relating driver and final position has an rsquare of 0.25 ... 

#### Uncertainty Quantification / Prediction Interval

If possible given your data, report all estimates with confidence / prediction / tolerance intervals. If not possible, clearly explain why it is not possible to provide intervals and document what sources of uncertainty are not quantified.


----------------------------------------

## Conclusion
testing testing testing

----------------------------------------

## Rubrics

Questions to answer:

- What question did you set out to answer?
- What data did you find to help answer that question?
- What is the relevant background on your question?
- What level of (quantified) certainty do you have in your results?
- What conclusions did you come to?
- What questions do you have remaining?
- Make sure your report contains at least one presentation-quality figure

Observed:

- (The usual stuff)
- Must provide background
- Must posit a question

Supported:

- (The usual stuff)
- Some analysis must support answering question

Assessed:

- (The usual stuff)
- All estimates must be provided with some quantification of uncertainty (e.g. confidence / prediction / tolerance intervals), OR a justification for why producing an interval is not possible and documentation for sources of uncertainty not accounted for.

Styled:

- (The usual stuff)
- Report must contain at least one presentation-quality figure

