---
title: "Modeling Avg Lap Time for Racers and Constructors"
output:
  github_document:
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(modelr)
```

```{r load data, include=FALSE}
df_avglaptime <- read_csv("processed_data/avglaptime.csv")
```

### EDA

```{r data checks, include = FALSE}
df_avglaptime %>% 
  glimpse()

df_avglaptime %>% 
  summary()
```

## Historgrams

```{r variable histgrams}
df_avglaptime %>% 
  filter(avg_lap <= 200000) %>% 
  ggplot(aes(avg_lap)) + 
  geom_histogram(bins = 50) +
  labs(
    title = "Average Lap Time Histogram", 
    x = "Average Lap Time (ms)"
  ) + 
  theme_minimal()

df_avglaptime %>% 
  ggplot(aes(circuit_avg_lap)) + 
  geom_histogram(bins = 60) + 
  labs(
    title = "Circuit Average Lap Time Histogram", 
    x = "Average Circuit Lap Time (ms)"
  ) + 
  theme_minimal()
```

```{r circuit average lap}
df_avglaptime %>% 
  mutate(
    year = as.factor(year), 
    circuit_name = fct_reorder(circuit_name, circuit_avg_lap)) %>% 
  ggplot(aes(circuit_name, circuit_avg_lap)) + 
  geom_jitter(aes(color = year), alpha = 1/14) +
  coord_flip() + 
  labs(title = "Circuit Average Lap for GPs containing Drivers Who Have Driven For Multiple Teams") +
  theme_minimal() +
  theme(plot.title = element_text(size = 9))
```
***Observations***

- Same Grands Prix can have vastly different times, most likely due to different courses under the same GP name
  - See United States GP and French GP
- Also the circuit average lap time seems to diverge onto one time and doesn't seem to diverge much
- `SO MUCH data` that it is hard to visualize the number of observations for a single GP 
  - What is the best way to visualize these observations??
  
### Modeling 

## Creating data frames

```{r data loading, include = FALSE}
df_stdavglap <- read_csv("processed_data/std_avg_laptime.csv")

# Creating row ID column for anti_join
df_stdavglap <- tibble::rowid_to_column(df_stdavglap, "ID")
```

## Visualizing Std Average Lap Time

```{r}
df_stdavglap %>% 
  ggplot(aes(std_avg_lap)) +
  geom_histogram(bins = 50)
```


## Creating Model

```{r modeling the entire data set}

fit_driver <-
  df_stdavglap %>%
  lm(
    data = .,
    formula = std_avg_lap ~ as.factor(driverId)
  )

rsquare(fit_driver, df_stdavglap)
mse(fit_driver, df_stdavglap)

fit_constructor <- 
  df_stdavglap %>% 
  lm(
    data = .,
    formula = std_avg_lap ~ as.factor(constructorId)
  )

rsquare(fit_constructor, df_stdavglap)
mse(fit_constructor, df_stdavglap)

fit_driver_constructor <- 
  df_stdavglap %>% 
  lm(
    data = .,
    formula = std_avg_lap ~ as.factor(driverId) + as.factor(constructorId)
  )

rsquare(fit_driver_constructor, df_stdavglap)
mse(fit_driver_constructor, df_stdavglap)

```


```{r train and validate data frames}
#Setting seed for temporary repeatability
set.seed("12345")

#Getting number of observations to feed into n for df_train
number_obs <- df_stdavglap %>% 
  group_by(driverId, circuitId) %>% 
  summarize(n = n()) %>% 
  pull(n)

#Sampling by 
df_train <-
  df_stdavglap %>%
  group_by(driverId, circuitId) %>% 
  slice_sample(n = 2) %>% 
  ungroup()

df_validate <-
  anti_join(
    df_stdavglap,
    df_train,
    by = "ID"
  )

df_train %>% arrange(ID)
df_validate %>% arrange(ID)
```



